---
title: "Movie Genre Overlappings, Ratings by Genre and Year"
author: "Ceshine Lee"
output:
  html_notebook
---

### (Exploring the Moives Dataset with rstudio/gt package)

Tables can be an effective way of communicating data. Though not as powerful in telling stories as charts, by cramming a lot of numbers into limited space, it can provide readers with accurate and potentially useful information which readers can interpret in their own ways.

I've come across this new R package [**gt**](https://github.com/rstudio/gt)(Easily generate information-rich, publication-quality tables from R) and decided to give it a try.

> With the gt package, anyone can make wonderful-looking tables using the R programming language.

Admittedly, the tables in the documents might not be the optimal way of presentation. They serve as a demonstration of what **gt** can do, and maybe also helpful enough for analyst in constructing their stories about this dataset.

(Dataset Source: [The Movies Dataset on Kaggle](https://www.kaggle.com/rounakbanik/the-movies-dataset).)

![](https://raw.githubusercontent.com/rstudio/gt/master/man/figures/gt_parts_of_a_table.svg?sanitize=true)

```{r, message=FALSE, results="hide"}
# install.packages("packrat")
# install.packages("pacman")
# install.packages("devtools")
# install.packages("tidyverse")
# devtools::install_github("rstudio/gt")
library(gt)
pacman::p_load("tidyverse")
pacman::p_load("data.table")
pacman::p_load("jsonlite")
pacman::p_load("lubridate")
```

## Read and clean the Metadata

Some of the fields read in here are not used in the later sections. However, they are kept as a reminder of the potentially useful information.

```{r read_clean_metadata}
metadata <- fread("data/movies_metadata.csv", select=c('adult', 'genres', 'release_date', 'original_language', 'original_title', 'id', 'imdb_id'), fill=T)
metadata <- metadata[!is.na(as.integer(id)) & (original_language == 'en') & (adult == "False"), ]
metadata[, id := as.integer(id)]
# Replace singles quote in 'genre' so 'fromJSON' function can parse it
metadata[, genre := gsub("\'","\"", metadata$genre)]
# Remove redundant fields
metadata[, c('original_language', 'adult', "imdb_id") := NULL]
```

## Parse Genres

Genre information is stored as JSON texts. We need to parse it and do some transformation.

```{r parse_genre}
genres <- metadata[, unlist(lapply(genre, fromJSON), recursive=F)['name'], by=id]
sorted.genres <- genres[, .N, by=name][order(-N)]
sorted.genres[1:20]
```

Assuming genres are assigned in order of representativeness, taking at most three top genres for each movie. Furthermore, only the top 6 genres are considered.

```{r}
# Dummy variable for the later dcast operation
genres[, dummy := 1]
# Only use the first 3 assigned genres
encoded.genres <- dcast(na.omit(genres[,.SD[1:3], by=id])[name %in% sorted.genres[1:6, name]],
                        id ~  name, value.var='dummy', fill=0)
data.w.genre <- merge(encoded.genres,
      metadata[, .(id, original_title, release_date)],
      all.x=T, by="id")
```

## Genre Statistics

Here I used a for loop to collect and calculate statstics genre by genre. I later figured out the better way to do this is probably using a *melt* call followed by several *ratio* column calculations. The later sections will use the *melt* approach instead.

```{r}
i <- 0
tmp.raw <- list()
tmp.ratio <- list()
for(name in sorted.genres[1:6, name]){
    i <- i + 1
    dt.tmp <- data.w.genre[get(name) == 1, lapply(.SD, sum), .SDcols=sorted.genres[1:6, name]]
    tmp.ratio[[i]] <- copy(dt.tmp)
    dt.tmp[, genre := name]
    tmp.raw[[i]] <- dt.tmp
    tmp.ratio[[i]] <- tmp.ratio[[i]] / dt.tmp[, get(name)]
    tmp.ratio[[i]][, genre := name]
}
genre.stats <- merge(rbindlist(tmp.raw), rbindlist(tmp.ratio), by="genre", suffixes=c("", ".ratio"), sort=F)
rm(tmp.raw, tmp.ratio, i)
```

The actual table declaration and rendering part. Unfortunately it requires a lot of manual labeling and assignments. But I think it is verbose and readable enough.

```{r}
gt_tbl <- gt(data = genre.stats, rowname_col = "genre")
gt_tbl %>%
  tab_header(
    title = "Movie Metadata Stats"
  ) %>%
  tab_spanner(
    label = "Genre Overlappings",
    columns = vars(Drama, Comedy, Thriller, Romance, Action, Horror)
  ) %>%
  tab_spanner(
    label = "Genre Overlappings(%)",
    columns = vars(Drama.ratio, Comedy.ratio, Thriller.ratio, Romance.ratio, Action.ratio, Horror.ratio)
  ) %>%
  fmt_percent(
    columns = vars(Drama.ratio, Comedy.ratio, Thriller.ratio, Romance.ratio, Action.ratio, Horror.ratio),
    decimals = 1,
    drop_trailing_zeros = F
  ) %>%
  tab_source_note(
    source_note = md("Source:  [\"The Movie Dataset\"](rounakbanik/the-movies-dataset) on Kaggle")
  ) %>%
  cols_label(
    Drama.ratio = "Drama",
    Comedy.ratio = "Comedy",
    Thriller.ratio = "Thriller",
    Romance.ratio = "Romance",
    Action.ratio = "Action",
    Horror.ratio = "Horror"
  ) %>%
  tab_stubhead_label(label = "Genre") %>%
  tab_style(
    style = cells_styles(
      text_size = px(12)),
    locations = list(
        cells_column_labels(columns = 1:12),
        cells_stub(),
        cells_data()
    )) %>%
  tab_style(
    style = cells_styles(
      text_decorate = "underline",
      text_weight = "bold"),
    locations = list(
        cells_data(columns=c(1), rows=c(1)),
        cells_data(columns=c(2), rows=c(2)),
        cells_data(columns=c(3), rows=c(3)),
        cells_data(columns=c(4), rows=c(4)),
        cells_data(columns=c(5), rows=c(5)),
        cells_data(columns=c(6), rows=c(6))
    )) %>%
  tab_style(
    style = cells_styles(
      text_color="lightgrey"),
    locations = list(
        cells_data(columns=c(7), rows=c(1)),
        cells_data(columns=c(8), rows=c(2)),
        cells_data(columns=c(9), rows=c(3)),
        cells_data(columns=c(10), rows=c(4)),
        cells_data(columns=c(11), rows=c(5)),
        cells_data(columns=c(12), rows=c(6))
    ))
```

*How to read the table*: For example, Drama (the first row) has 13008 movies (the first column), and 2639 of them are also under Comedy, which is 20.3% of the 13008 movies.

The diagonal elements of 'Genre Overlappings' section are emphasized because they represent the movie count of the respective genre. Similarily, the diagonal elements of 'Genre Overlappings(%)' section are somewhat hiddened since they convey almost no useful information.


## Reading Ratings

Now we turn our attention to movie ratings (from movielens). We only consider the average rating of a movie (which of course is not perfect).

```{r, read_ratings}
ratings <- fread("data/ratings.csv", select=c('movieId', 'rating'))
ratings.stats <- ratings[, .(n_ratings=.N, avg_rating=mean(rating)), by=movieId]
rm(ratings)
head(ratings.stats)
```

Here we remove movies with less than 500 ratings.

```{r}
data.w.genre.ratings <- merge(data.w.genre, ratings.stats, by.y="movieId", by.x="id")
# Only keep moives with >= 500 ratings
data.w.genre.ratings <- data.w.genre.ratings[n_ratings >= 500]
head(data.w.genre.ratings)
```

## Rating statistics by Genre

Using a *melt* call followed by a bunch of data.table operations to prepare the metrics:

```{r}
rating.by.genre <- melt(
    data.w.genre.ratings, id=c("id", "avg_rating", "n_ratings"), measure.vars=c("Action", "Comedy", "Drama", "Horror", "Romance", "Thriller"), variable.name="genre"
    )[value==1][
    , .(n_movies=.N, min_rating=min(avg_rating), avg_rating=mean(avg_rating), med_rating=median(avg_rating), max_rating=max(avg_rating), sd_rating=sd(avg_rating), n_ratings=sum(n_ratings), avg_n_ratings=mean(n_ratings), max_n_ratings=max(n_ratings), min_n_ratings=min(n_ratings)), by=genre]
head(rating.by.genre)
```

Follows roughly the same recipe (unfortunately I did not find a reasonable way to make use of row group feature of *gt* for this dataset):

```{r}
gt_tbl <- gt(data = rating.by.genre, rowname_col = "genre")
gt_tbl %>%
  tab_header(
    title = "Movie Average Ratings by Genre",
    subtitle = "with # of Ratings >= 500"
  ) %>%
  tab_spanner(
    label = "Average Rating (1-5)",
    columns = vars(min_rating, avg_rating, med_rating, max_rating, sd_rating)
  ) %>%
  fmt_number(
    columns = vars(min_rating, avg_rating, med_rating, max_rating, sd_rating),
    decimals = 2,
    drop_trailing_zeros = F
  ) %>%
  tab_spanner(
    label = "# of Ratings",
    columns = vars(n_ratings, min_n_ratings, avg_n_ratings, max_n_ratings)
  ) %>%
  fmt_number(
    columns = vars(n_ratings, min_n_ratings, avg_n_ratings, max_n_ratings),
    decimals = 0,
    drop_trailing_zeros = F
  ) %>%
  tab_source_note(
    source_note = md("Source:  [\"The Movie Dataset\"](rounakbanik/the-movies-dataset) on Kaggle")
  ) %>%
  cols_label(
    min_rating = "Min",
    avg_rating = "Avg",
    med_rating = "Med",
    max_rating = "Max",
    n_ratings = "Total",
    min_n_ratings = "Min",
    avg_n_ratings = "Avg",
    max_n_ratings = "Max",
    n_movies="Movies",
    sd_rating="Stdev"
  ) %>%
  tab_stubhead_label(label = "Genre") %>%
  tab_style(
    style = cells_styles(
      text_size = px(14)),
    locations = list(
        cells_column_labels(columns = 1:10),
        cells_stub(),
        cells_data()
    ))
```

Suprisingly, the distribution of ratings are quite similar across all genres. Maybe movielens has done some normalization on the ratings?


## Rating statistics by Genre and Year

We'd also like to know if the distributions of ratings change over time. First we plot the histogram of the years in which the movies were released. This is a case where a histogram is far more readable than a table.

```{r}
data.w.genre.ratings[,release_year:=year(ymd(release_date))]
ggplot(data.w.genre.ratings[release_year >= 1950][order(release_year),.N, by=release_year], aes(x=release_year, y=N)) +
    geom_bar(stat="identity") + ggtitle("# of Movie by Year") + theme_bw() + scale_x_continuous(breaks=seq(1950, 2020, 10))
```

Now we only take movies released after 1979 and before 2010 into account, and put them in to buckets each representing a decade.

```{r}
data.w.genre.ratings <- data.w.genre.ratings[(release_year >= 1980) & (release_year < 2010)]
data.w.genre.ratings[, release_decade := (release_year %/% 10) * 10 ]
```

The same old *melt* trick:

```{r}
rating.by.genre.year <- melt(
    data.w.genre.ratings, id=c("id", "avg_rating", "n_ratings", "release_decade"), measure.vars=c("Action", "Comedy", "Drama", "Horror", "Romance", "Thriller"), variable.name="genre"
    )[value==1][
    , .(n_movies=.N, min_rating=min(avg_rating), avg_rating=mean(avg_rating), med_rating=median(avg_rating), max_rating=max(avg_rating), sd_rating=sd(avg_rating), n_ratings=sum(n_ratings), avg_n_ratings=mean(n_ratings), max_n_ratings=max(n_ratings), min_n_ratings=min(n_ratings)), by=.(genre, release_decade)]
```

Even more manual labeling and a *dcast* call to handle the extra *decade* dimension:

```{r}
gt_tbl <- gt(
    data = dcast(rating.by.genre.year, genre ~ release_decade, value.var = c("avg_rating", "sd_rating", "avg_n_ratings", "n_movies"))[, .(genre,
        n_movies_1980, avg_rating_1980, sd_rating_1980, avg_n_ratings_1980,
        n_movies_1990, avg_rating_1990, sd_rating_1990, avg_n_ratings_1990,
        n_movies_2000, avg_rating_2000, sd_rating_2000, avg_n_ratings_2000
    )],
    rowname_col = "genre")
gt_tbl %>%
  tab_header(
    title = "Movie Average Ratings by Genre & Year",
    subtitle = "with # of Ratings >= 500"
  ) %>%
  tab_spanner(
    label = "1980s",
    columns = vars(n_movies_1980, avg_rating_1980, sd_rating_1980, avg_n_ratings_1980)
  ) %>%
  tab_spanner(
    label = "1990s",
    columns = vars(n_movies_1990, avg_rating_1990, sd_rating_1990, avg_n_ratings_1990)
  ) %>%
  tab_spanner(
    label = "2000s",
    columns = vars(n_movies_2000, avg_rating_2000, sd_rating_2000, avg_n_ratings_2000)
  ) %>%
  fmt_number(
    columns = vars(avg_rating_1980, sd_rating_1980, avg_rating_1990, sd_rating_1990, avg_rating_2000, sd_rating_2000),
    decimals = 2,
    drop_trailing_zeros = F
  ) %>%
  fmt_number(
    columns = vars(avg_n_ratings_1980, avg_n_ratings_1990, avg_n_ratings_2000, n_movies_1980, n_movies_1990, n_movies_2000),
    decimals = 0,
    drop_trailing_zeros = F
  ) %>%
  tab_source_note(
    source_note = md("Source:  [\"The Movie Dataset\"](rounakbanik/the-movies-dataset) on Kaggle")
  ) %>%
  tab_footnote(
    footnote = "#: Number of movies",
    cells_column_labels(columns = c(1, 5, 9))
  ) %>%
  tab_footnote(
    footnote = "Avg #: Average number of ratings.",
    cells_column_labels(columns = c(4, 8, 12))
  ) %>%
  cols_label(
    n_movies_1980 = "#",
    n_movies_1990 = "#",
    n_movies_2000 = "#",
    avg_rating_1980 = "Avg",
    avg_rating_1990 = "Avg",
    avg_rating_2000 = "Avg",
    sd_rating_1980 = "Stdev",
    sd_rating_1990 = "Stdev",
    sd_rating_2000 = "Stdev",
    avg_n_ratings_1980 = "Avg #",
    avg_n_ratings_1990 = "Avg #",
    avg_n_ratings_2000 = "Avg #"
  ) %>%
  tab_stubhead_label(label = "Genre") %>%
  tab_style(
    style = cells_styles(
      text_size = px(14)),
    locations = list(
        cells_column_labels(columns = 1:12),
        cells_stub(),
        cells_data()
    ))
```

It appears the variances of ratings in 1980s are lower than in 1990s and 2000s. (A boxplot or scatterplot might be more appropriate here.)

## Acknowledgements

* Erik Bruin's Kaggle notebook ["Movie recommendation systems based on TMDB"](https://www.kaggle.com/erikbruin/movie-recommendation-systems-based-on-tmdb) has been a great source of ideas r.t. manipulating this dataset using R.

