---
title: "Movie Genre Overlappings, Ratings by Genre and Year"
output:
  html_notebook
---

# Exploring the Moives Dataset with rstudio/gt package

Dataset Source: [The Movies Dataset on Kaggle](https://www.kaggle.com/rounakbanik/the-movies-dataset).

```{r, message=FALSE, results="hide"}
# install.packages("packrat")
# install.packages("pacman")
# install.packages("devtools")
# install.packages("tidyverse")
# devtools::install_github("rstudio/gt")
library(gt)
pacman::p_load("tidyverse")
pacman::p_load("data.table")
pacman::p_load("jsonlite")
pacman::p_load("lubridate")
```

## Read and clean the Metadata
```{r read_clean_metadata}
metadata <- fread("data/movies_metadata.csv", select=c('adult', 'genres', 'release_date', 'original_language', 'original_title', 'id', 'imdb_id'), fill=T)
metadata <- metadata[!is.na(as.integer(id)) & (original_language == 'en') & (adult == "False"), ]
metadata[, id := as.integer(id)]
metadata[, genre := gsub("\'","\"", metadata$genre)]
metadata[, c('original_language', 'adult') := NULL]
```

## Parse Genre
```{r parse_genre}
genres <- metadata[, unlist(lapply(genre, fromJSON), recursive=F)['name'], by=id]
sorted.genres <- genres[, .N, by=name][order(-N)]
sorted.genres[1:20]
```

```{r}
genres[, dummy := 1]
# Only use the first 3 assigned genres 
encoded.genres <- dcast(na.omit(genres[,.SD[1:3], by=id])[name %in% sorted.genres[1:6, name]], 
                        id ~  name, value.var='dummy', fill=0)
data.w.genre <- merge(encoded.genres,
      metadata[, .(id, original_title, imdb_id, release_date)], 
      all.x=T, by="id")
```

## Genre Statistics
```{r}
i <- 0
tmp.raw <- list() 
tmp.ratio <- list()
for(name in sorted.genres[1:6, name]){
    i <- i + 1
    dt.tmp <- data.w.genre[get(name) == 1, lapply(.SD, sum), .SDcols=sorted.genres[1:6, name]]
    tmp.ratio[[i]] <- copy(dt.tmp)
    dt.tmp[, genre := name]
    tmp.raw[[i]] <- dt.tmp
    tmp.ratio[[i]] <- tmp.ratio[[i]] / dt.tmp[, get(name)]
    tmp.ratio[[i]][, genre := name]
}
genre.stats <- merge(rbindlist(tmp.raw), rbindlist(tmp.ratio), by="genre", suffixes=c("", ".ratio"), sort=F)
rm(tmp.raw, tmp.ratio, i)
```


```{r}
gt_tbl <- gt(data = genre.stats, rowname_col = "genre")
gt_tbl %>%
  tab_header(
    title = "Movie Metadata Stats"
  ) %>%
  tab_spanner(
    label = "Genre Overlappings",
    columns = vars(Drama, Comedy, Thriller, Romance, Action, Horror)
  ) %>%
  tab_spanner(
    label = "Genre Overlappings(%)",
    columns = vars(Drama.ratio, Comedy.ratio, Thriller.ratio, Romance.ratio, Action.ratio, Horror.ratio)
  ) %>%
  fmt_percent(
    columns = vars(Drama.ratio, Comedy.ratio, Thriller.ratio, Romance.ratio, Action.ratio, Horror.ratio),
    decimals = 1,
    drop_trailing_zeros = F
  ) %>%   
  tab_source_note(
    source_note = md("Source:  [\"The Movie Dataset\"](rounakbanik/the-movies-dataset) on Kaggle")
  ) %>%
  cols_label(
    Drama.ratio = "Drama",
    Comedy.ratio = "Comedy",
    Thriller.ratio = "Thriller",
    Romance.ratio = "Romance", 
    Action.ratio = "Action", 
    Horror.ratio = "Horror"
  ) %>%   
  tab_stubhead_label(label = "Genre") %>%
  tab_style(
    style = cells_styles(
      text_size = px(12)),
    locations = list(
        cells_column_labels(columns = 1:12), 
        cells_stub(),
        cells_data()
    )) %>%   
  tab_style(
    style = cells_styles(
      text_decorate = "underline",
      text_weight = "bold"),
    locations = list(
        cells_data(columns=c(1), rows=c(1)),
        cells_data(columns=c(2), rows=c(2)),
        cells_data(columns=c(3), rows=c(3)),
        cells_data(columns=c(4), rows=c(4)),
        cells_data(columns=c(5), rows=c(5)),
        cells_data(columns=c(6), rows=c(6))
    )) %>%   
  tab_style(
    style = cells_styles(
      text_color="lightgrey"),
    locations = list(
        cells_data(columns=c(7), rows=c(1)),
        cells_data(columns=c(8), rows=c(2)),
        cells_data(columns=c(9), rows=c(3)),
        cells_data(columns=c(10), rows=c(4)),
        cells_data(columns=c(11), rows=c(5)),
        cells_data(columns=c(12), rows=c(6))
    ))
```

## Reading Ratings
```{r, read_ratings}
ratings <- fread("data/ratings.csv", select=c('movieId', 'rating'))
ratings.stats <- ratings[, .(n_ratings=.N, avg_rating=mean(rating)), by=movieId]
rm(ratings)
head(ratings.stats)
```

```{r}
data.w.genre.ratings <- merge(data.w.genre, ratings.stats, by.y="movieId", by.x="id")
# Only keep moives with >= 500 ratings
data.w.genre.ratings <- data.w.genre.ratings[n_ratings >= 500]
head(data.w.genre.ratings)
```

## Rating statistics by Genre

```{r}
rating.by.genre <- melt(
    data.w.genre.ratings, id=c("id", "avg_rating", "n_ratings"), measure.vars=c("Action", "Comedy", "Drama", "Horror", "Romance", "Thriller"), variable.name="genre"
    )[value==1][
    , .(n_movies=.N, min_rating=min(avg_rating), avg_rating=mean(avg_rating), med_rating=median(avg_rating), max_rating=max(avg_rating), sd_rating=sd(avg_rating), n_ratings=sum(n_ratings), avg_n_ratings=mean(n_ratings), max_n_ratings=max(n_ratings), min_n_ratings=min(n_ratings)), by=genre]
head(rating.by.genre)
```

```{r}
gt_tbl <- gt(data = rating.by.genre, rowname_col = "genre")
gt_tbl %>%
  tab_header(
    title = "Movie Average Ratings by Genre",
    subtitle = "with # of Ratings >= 500"
  ) %>%
  tab_spanner(
    label = "Average Rating (1-5)",
    columns = vars(min_rating, avg_rating, med_rating, max_rating, sd_rating)
  ) %>%
  fmt_number(
    columns = vars(min_rating, avg_rating, med_rating, max_rating, sd_rating),
    decimals = 2,
    drop_trailing_zeros = F
  ) %>%       
  tab_spanner(
    label = "# of Ratings",
    columns = vars(n_ratings, min_n_ratings, avg_n_ratings, max_n_ratings)
  ) %>%
  fmt_number(
    columns = vars(n_ratings, min_n_ratings, avg_n_ratings, max_n_ratings),
    decimals = 0,
    drop_trailing_zeros = F
  ) %>%    
  tab_source_note(
    source_note = md("Source:  [\"The Movie Dataset\"](rounakbanik/the-movies-dataset) on Kaggle")
  ) %>%
  cols_label(
    min_rating = "Min", 
    avg_rating = "Avg", 
    med_rating = "Med",
    max_rating = "Max",
    n_ratings = "Total", 
    min_n_ratings = "Min", 
    avg_n_ratings = "Avg",
    max_n_ratings = "Max",
    n_movies="Movies",
    sd_rating="Stdev"
  ) %>%   
  tab_stubhead_label(label = "Genre") %>%
  tab_style(
    style = cells_styles(
      text_size = px(14)),
    locations = list(
        cells_column_labels(columns = 1:10), 
        cells_stub(),
        cells_data()
    ))
```

## Rating statistics by Genre and Year

```{r}
data.w.genre.ratings[,release_year:=year(ymd(release_date))]
ggplot(data.w.genre.ratings[release_year >= 1950][order(release_year),.N, by=release_year], aes(x=release_year, y=N)) + 
    geom_bar(stat="identity") + ggtitle("# of Movie by Year") + theme_bw() + scale_x_continuous(breaks=seq(1950, 2020, 10))
```

```{r}
data.w.genre.ratings <- data.w.genre.ratings[(release_year >= 1980) & (release_year < 2010)]
data.w.genre.ratings[, release_decade := (release_year %/% 10) * 10 ]
```


```{r}
rating.by.genre.year <- melt(
    data.w.genre.ratings, id=c("id", "avg_rating", "n_ratings", "release_decade"), measure.vars=c("Action", "Comedy", "Drama", "Horror", "Romance", "Thriller"), variable.name="genre"
    )[value==1][
    , .(n_movies=.N, min_rating=min(avg_rating), avg_rating=mean(avg_rating), med_rating=median(avg_rating), max_rating=max(avg_rating), sd_rating=sd(avg_rating), n_ratings=sum(n_ratings), avg_n_ratings=mean(n_ratings), max_n_ratings=max(n_ratings), min_n_ratings=min(n_ratings)), by=.(genre, release_decade)]
head(rating.by.genre.year)
```

```{r}
gt_tbl <- gt(
    data = dcast(rating.by.genre.year, genre ~ release_decade, value.var = c("avg_rating", "sd_rating", "avg_n_ratings", "n_movies"))[, .(genre,
        n_movies_1980, avg_rating_1980, sd_rating_1980, avg_n_ratings_1980, 
        n_movies_1990, avg_rating_1990, sd_rating_1990, avg_n_ratings_1990,
        n_movies_2000, avg_rating_2000, sd_rating_2000, avg_n_ratings_2000
    )], 
    rowname_col = "genre")
gt_tbl %>%
  tab_header(
    title = "Movie Average Ratings by Genre & Year",
    subtitle = "with # of Ratings >= 500"
  ) %>%
  tab_spanner(
    label = "1980s",
    columns = vars(n_movies_1980, avg_rating_1980, sd_rating_1980, avg_n_ratings_1980)
  ) %>%
  tab_spanner(
    label = "1990s",
    columns = vars(n_movies_1990, avg_rating_1990, sd_rating_1990, avg_n_ratings_1990)
  ) %>%
  tab_spanner(
    label = "2000s",
    columns = vars(n_movies_2000, avg_rating_2000, sd_rating_2000, avg_n_ratings_2000)
  ) %>%
  fmt_number(
    columns = vars(avg_rating_1980, sd_rating_1980, avg_rating_1990, sd_rating_1990, avg_rating_2000, sd_rating_2000),
    decimals = 2,
    drop_trailing_zeros = F
  ) %>%       
  fmt_number(
    columns = vars(avg_n_ratings_1980, avg_n_ratings_1990, avg_n_ratings_2000, n_movies_1980, n_movies_1990, n_movies_2000),
    decimals = 0,
    drop_trailing_zeros = F
  ) %>%    
  tab_source_note(
    source_note = md("Source:  [\"The Movie Dataset\"](rounakbanik/the-movies-dataset) on Kaggle")
  ) %>%
  tab_footnote(
    footnote = "#: Number of movies",
    cells_column_labels(columns = c(1, 5, 9))
  ) %>%
  tab_footnote(
    footnote = "Avg #: Average number of ratings.",
    cells_column_labels(columns = c(4, 8, 12))
  ) %>%    
  cols_label(
    n_movies_1980 = "#",
    n_movies_1990 = "#",
    n_movies_2000 = "#",
    avg_rating_1980 = "Avg", 
    avg_rating_1990 = "Avg", 
    avg_rating_2000 = "Avg", 
    sd_rating_1980 = "Stdev", 
    sd_rating_1990 = "Stdev",
    sd_rating_2000 = "Stdev",
    avg_n_ratings_1980 = "Avg #", 
    avg_n_ratings_1990 = "Avg #", 
    avg_n_ratings_2000 = "Avg #"
  ) %>%   
  tab_stubhead_label(label = "Genre") %>%
  tab_style(
    style = cells_styles(
      text_size = px(14)),
    locations = list(
        cells_column_labels(columns = 1:12), 
        cells_stub(),
        cells_data()
    ))
```

## Acknowledgements

* Erik Bruin's Kaggle notebook ["Movie recommendation systems based on TMDB"](https://www.kaggle.com/erikbruin/movie-recommendation-systems-based-on-tmdb) has been a great source of ideas r.t. manipulating this dataset using R.

